---
title: "Oregon Labor Day Fires"
author: "Mazama Science"
date: "2020-09-18"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 9)
```

## Interactive visualizations of PM2.5 monitoring data

This web page demonstrates a new **tiotemp** package being developed by 
Mazama Science to provide interactive versions of common data visualizations
for air quality monitoring data.

```{r echo = FALSE, warning = FALSE, message = FALSE}
library(tiotemp)
library(PWFSLSmoke)
```

We will use data obtained with the **PWFSLSmoke** monitoring package for Oregon
and Washginton covering the Oregon fires which began on Labor Day weekend, 2020.

```{r create_orwa}
# Get 2020 data
orwa_latest <-
  monitor_loadLatest() %>%
  monitor_subsetBy(timezone == "America/Los_Angeles") %>%
  monitor_subsetBy(stateCode %in% c("OR","WA"))
  
orwa_2020 <- 
  monitor_load(20200101, 20201001, monitorIDs = orwa_latest$meta$monitorID)

# Get annual data but use the IDs to speed things up
orwa_summer <-
  orwa_2020 %>%
  monitor_subset(tlim = c(20200701, 20201101))

# Get the fire/smoke days
orwa_smoke <-
  orwa_summer %>%
  monitor_subset(tlim = c(20200907, 20200914))
```

For a little context for how terrible the smoke event was, the following graphic
plots the values for 132 air quality monitors in Washington and Oregon. The date
typically associated with the worst air quality of the year -- early July 5 --
pales in comparisin to the smoke monster of mid September.

```{r set_par, echo = FALSE}
par(mar = c(5,4,6,2) + 0.1)
```

```{r orwa_summer_timeseriesPlot}
# Create the basic plot
monitor_timeseriesPlot(
  orwa_summer,
  style = 'gnats',
  ylim = c(0,500),
  xpd = NA
)

# Add annotations
addAQILegend("top")
addAQIStackedBar()
addAQILines()
title("Oregon/Washington Air Quality in 2020 -- the September of Smoke")
```

```{r restore_par, echo = FALSE}
par(mar = c(5,4,4,2) + 0.1)
```

### TimeseriesMap

The first interactive visualization is a timeseries map of hourly data during the first
week of the smoke. This is a leaflet map with a built in slider that allows you 
to scroll through time or just watch things evolve by hitting the play button. 

```{r orwa_timeseriesMap}
timeseriesMap(
  data = orwa_smoke$data,
  meta = orwa_smoke$meta,
  index = "monitorID",
  label = "siteName",
  colors = c("#00E400","#FFFF00","#FF7E00","#FF0000","#8F3F97","#7E0023"),
  breaks = c(12.0, 35.5,  55.5, 150.5, 250.5)
)
```

### TimeseriesBarchart

The seond interactive visualization is a timeseries barchart that allows you to "brush"
over the bar chart to zoom in or double click to zoom out.

For this chart, the data is from a monitor is in downtown Seattle.

```{r orwa_timeseriesBarchart}
seattle <- monitor_subset(orwa_summer, monitorIDs = "530330030_01")
timeseriesBarChart(
  seattle$data, 
  seattle$meta,
  index = "monitorID",
  label = "siteName",
  colors = c("#00E400","#FFFF00","#FF7E00","#FF0000","#8F3F97","#7E0023"),
  breaks = c(12.0, 35.5,  55.5, 150.5, 250.5), 
  ymax = 200
)
```

### TimeseriesCalendar

The third visualizaiton is a calendar plot which displays daily averages for 
a subset of months or for an entire year.

Data from the same downtown Seattle monitor is used in this chart. July 5'th 
barely shows up while June 25'th was the date of a large building fire near the
monitor. The September smoke stands out.

```{r orwa_timeseriesCalendar}
seattle <- monitor_subset(orwa_2020, monitorIDs = "530330030_01")
timeseriesCalendar(
  seattle$data, 
  seattle$meta,
  index = "monitorID",
  label = "siteName",
  full = FALSE,
  colors = c("#00E400","#FFFF00","#FF7E00","#FF0000","#8F3F97","#7E0023"),
  breaks = c(12.0, 35.5,  55.5, 150.5, 250.5)
)
```
